---
title: "caret package take home"
author: "lianrui"
date: "2017年9月30日"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

## Visualizations
```{r, fig.width=12, fig.height=12}
data(iris)
library(AppliedPredictiveModeling)
transparentTheme(trans = .4)
library(caret)
featurePlot(x = iris[, 1:4], 
            y = iris$Species, 
            plot = "pairs",
            ## Add a key at the top
            auto.key = list(columns = 3))
```

```{r, fig.height=12, fig.width=12}
featurePlot(x = iris[, 1:4], 
            y = iris$Species, 
            plot = "ellipse",
            ## Add a key at the top
            auto.key = list(columns = 3))
```


```{r}
transparentTheme(trans = .9)
featurePlot(x = iris[, 1:4], 
            y = iris$Species,
            plot = "density", 
            ## Pass in options to xyplot() to 
            ## make it prettier
            scales = list(x = list(relation="free"), 
                          y = list(relation="free")), 
            adjust = 1.5, 
            pch = "|", 
            layout = c(4, 1), 
            auto.key = list(columns = 3))
```

```{r}
featurePlot(x = iris[, 1:4], 
            y = iris$Species, 
            plot = "box", 
            ## Pass in options to bwplot() 
            scales = list(y = list(relation="free"),
                          x = list(rot = 90)),  
            layout = c(4,1 ), 
            auto.key = list(columns = 2))
```

```{r}

```

## preprocessing
### Creating dummy variables
Two approaches of creating dummy variables for factors:
* 1. model.matrix in R base
* 2. dummyVars in caret package

In model.matrix, the first variable is taken as intercept. 
In contrast, dummyVars will return just a matrix, No intercept is returns, this is not useful some models requiring dependents and predicters, e.g., lm. 

```{r}
library(earth)
data(etitanic)
head(etitanic)
str(etitanic)
# etitanic has two factor variable, pclass, sex
model.matrix(survived ~ ., data = etitanic)
?model.matrix

head(model.matrix(survived ~ ., data = etitanic))
```


```{r}
dummies <- dummyVars(survived ~ ., data = etitanic)
head(dummies)
head(predict(dummies, newdata = etitanic))
```


### Zero- and Near Zero-Variance Predictors

example
```{r}
data(mdrr)
nzv <- nearZeroVar(mdrrDescr, saveMetrics = TRUE)
str(nzv)
nzv[nzv$nzv,][1:10,]
```

