---
title       : Introduction to the R Language
subtitle    : Data Types and Basic Operations
author      : Roger Peng, Associate Professor
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

## Objects

R has five basic or “atomic” classes of objects:

-   character

-   numeric (real numbers)

-   integer

-   complex

-   logical (True/False)

The most basic object is a vector

-   A vector can only contain objects of the same class

-   BUT: The one exception is a *list*, which is represented as a vector
    but can contain objects of different classes (indeed, that’s usually
    why we use them)

Empty vectors can be created with the ```vector()``` function.
```{r}
vector()
```


---

## Numbers

-   Numbers in R a generally treated as numeric objects (i.e. double
    precision real numbers)

-   If you explicitly want an integer, you need to specify the ```L```
    suffix
    
```{r}
a <- 3L; b <- 3
class(a);class(b)
```

-   Ex: Entering ```1``` gives you a numeric object; entering ```1L```
    explicitly gives you an integer.

-   There is also a special number ```Inf``` which represents infinity;
    e.g. ```1 / 0```; ```Inf``` can be used in ordinary calculations;
    e.g. ```1 / Inf``` is 0
    
```{r}
1/0; 0/0
```

-   The value ```NaN``` represents an undefined value (“not a number”);
    e.g. 0 / 0; ```NaN``` can also be thought of as a missing value
    (more on that later)

---

## Attributes

R objects can have attributes

-   names, dimnames

-   dimensions (e.g. matrices, arrays)

-   class

-   length

-   other user-defined attributes/metadata

Attributes of an object can be accessed using the ```attributes()```
function.
```{r}
data("mtcars")
attributes(mtcars)
str(mtcars)
```

---

## Entering Input

At the R prompt we type expressions. The `<-` symbol is the assignment operator.

```{r}
x <- 1
print(x)
x
msg <- "hello"
msg
```

The grammar of the language determines whether an expression is complete or not.

```{r}
x <-  ## Incomplete expression
```

The # character indicates a comment. Anything to the right of the # (including the # itself) is ignored.

---

## Evaluation

When a complete expression is entered at the prompt, it is evaluated and the result of the evaluated expression is returned. The result may be auto-printed.

```{r}
x <- 5  ## nothing printed
x       ## auto-printing occurs
print(x)  ## explicit printing
```

The `[1]` indicates that `x` is a vector and `5` is the first element.

---

## Printing

```{r}
x <- 1:20 
x
```

The `:` operator is used to create integer sequences.

---

## Creating Vectors

The `c()` function can be used to create vectors of objects.

```{r}
x <- c(0.5, 0.6)       ## numeric
x <- c(TRUE, FALSE)    ## logical
x <- c(T, F)           ## logical
x <- c("a", "b", "c")  ## character
x <- 9:29              ## integer
x <- c(1+0i, 2+4i)     ## complex
```

Using the `vector()` function

```{r}
x <- vector("numeric", length = 10) 
x
```

---

## Mixing Objects

What about the following?

```r
> y <- c(1.7, "a")   ## character
> y <- c(TRUE, 2)    ## numeric
> y <- c("a", TRUE)  ## character
```

When different objects are mixed in a vector, _coercion_ occurs so that every element in the vector is of the same class.

---

## Explicit Coercion

Objects can be explicitly coerced from one class to another using the `as.*` functions, if available.

```{r}
x <- 0:6
class(x)

as.numeric(x)

as.logical(x)

as.character(x)

```

---

## Explicit Coercion

Nonsensical coercion results in `NA`s.

```r
> x <- c("a", "b", "c")
> as.numeric(x)
[1] NA NA NA
Warning message:
NAs introduced by coercion
> as.logical(x)
[1] NA NA NA
> as.complex(x)
[1] NA NA NA
Warning message:
NAs introduced by coercion 
```

---

## Matrices

Matrices are vectors with a _dimension_ attribute. The dimension attribute is itself an integer vector of length 2 (nrow, ncol)

```r
> m <- matrix(nrow = 2, ncol = 3) 
> m
     [,1] [,2] [,3]
[1,]   NA   NA   NA
[2,]   NA   NA   NA
> dim(m)
[1] 2 3
> attributes(m)
$dim
[1] 2 3
```

---

## Matrices (cont’d)

Matrices are constructed _column-wise_, so entries can be thought of starting in the “upper left” corner and running down the columns.

```r
> m <- matrix(1:6, nrow = 2, ncol = 3) 
> m
     [,1] [,2] [,3]
[1,]    1    3    5
[2,]    2    4    6
```

---

## Matrices (cont’d)

Matrices can also be created directly from vectors by adding a dimension attribute.

```r
> m <- 1:10 
> m
[1] 1 2 3 4 5 6 7 8 9 10 
> dim(m) <- c(2, 5)
> m
     [,1] [,2] [,3] [,4] [,5]
[1,]    1    3    5    7    9
[2,]    2    4    6    8   10
```

---

## cbind-ing and rbind-ing

Matrices can be created by _column-binding_ or _row-binding_ with `cbind()` and `rbind()`.

```r
> x <- 1:3
> y <- 10:12
> cbind(x, y)
     x  y 
[1,] 1 10 
[2,] 2 11 
[3,] 3 12
> rbind(x, y) 
  [,1] [,2] [,3]
x    1    2    3
y   10   11   12
```

---

## Lists

Lists are a special type of vector that can contain elements of different classes. Lists are a very important data type in R and you should get to know them well.

```r
> x <- list(1, "a", TRUE, 1 + 4i) 
> x
[[1]]
[1] 1

[[2]] 
[1] "a"

[[3]]
[1] TRUE

[[4]]
[1] 1+4i
```

---

## Factors

Factors are used to represent categorical data. Factors can be unordered or ordered. One can think of a factor as an integer vector where each integer has a _label_.

- Factors are treated specially by modelling functions like `lm()` and `glm()`

- Using factors with labels is _better_ than using integers because factors are self-describing; having a variable that has values “Male” and “Female” is better than a variable that has values 1 and 2.

---

## Factors

```r
> x <- factor(c("yes", "yes", "no", "yes", "no")) 
> x
[1] yes yes no yes no
Levels: no yes
> table(x) 
x
no yes 
 2   3
> unclass(x)
[1] 2 2 1 2 1
attr(,"levels")
[1] "no"  "yes"
```

---

## Factors

The order of the levels can be set using the `levels` argument to `factor()`. This can be important in linear modelling because the first level is used as the baseline level.


```r
> x <- factor(c("yes", "yes", "no", "yes", "no"),
              levels = c("yes", "no"))
> x
[1] yes yes no yes no 
Levels: yes no
```

---

## Missing Values

Missing values are denoted by `NA` or `NaN` for undefined mathematical operations. 

- `is.na()` is used to test objects if they are `NA`

- `is.nan()` is used to test for `NaN`

- `NA` values have a class also, so there are integer `NA`, character `NA`, etc.

- A `NaN` value is also `NA` but the converse is not true

---

## Missing Values

```{r}
x <- c(1, 2, NA, NA, NA)
y <- c(NA, NA, 1, 2, 3)
d <- data.frame(x = x, y = y)
## the number of missing values. 
b <- apply(d, 2, function(x)sum(is.na(x)))
## identify the index of missing values in a data frame
b <- apply(d, 2, function(x)which(is.na(x)))

library(reshape2)
dmelt <- melt(d)
dmelt
library(ggplot2)
g <- ggplot(data = dmelt, mapping = aes(x = variable, y = is.na(value)+value))
g <- g + geom_point()
g

#return the position of na in a vector r
a <- which(is.na(x))

is.nan(x)

x <- c(1, 2, NaN, NA, 4)
is.na(x)
is.nan(x)

```

---

## Data Frames

Data frames are used to store tabular data

- They are represented as a special type of list where every element of the list has to have the same length

- Each element of the list can be thought of as a column and the length of each element of the list is the number of rows

- Unlike matrices, data frames can store different classes of objects in each column (just like lists); matrices must have every element be the same class

- Data frames also have a special attribute called `row.names`

- Data frames are usually created by calling `read.table()` or `read.csv()`

- Can be converted to a matrix by calling `data.matrix()`

---

## Data Frames

```r
> x <- data.frame(foo = 1:4, bar = c(T, T, F, F)) 
> x
  foo   bar
1   1  TRUE
2   2  TRUE
3   3 FALSE
4   4 FALSE
> nrow(x)
[1] 4
> ncol(x)
[1] 2
```

---

## Names

R objects can also have names, which is very useful for writing readable code and self-describing objects.

```r
> x <- 1:3
> names(x)
NULL
> names(x) <- c("foo", "bar", "norf") 
> x
foo bar norf 
  1   2    3
> names(x)
[1] "foo"  "bar"  "norf"
```

---

## Names

Lists can also have names.

```r
> x <- list(a = 1, b = 2, c = 3) 
> x
$a
[1] 1

$b 
[1] 2

$c 
[1] 3
```

---

## Names

And matrices.

```r
> m <- matrix(1:4, nrow = 2, ncol = 2)
> dimnames(m) <- list(c("a", "b"), c("c", "d")) 
> m
  c d 
a 1 3 
b 2 4
```

---

## Summary

Data Types

- atomic classes: numeric, logical, character, integer, complex \

- vectors, lists

- factors

- missing values

- data frames

- names
